---
title: "hbw_DF_DSM"
author: "Abigail Schiffmiller"
date: "2025-06-23"
output:
  pdf_document: default
  html_document: default
---

## libraries

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is based off the Dolphin dsm example from the distance sampling website: [`http://examples.distancesampling.org/dsm-line-dolphins/mexico-analysis.html`]

```{r libraries , eval=FALSE, include= TRUE}
#install.packages(c("dsm", "Distance", "distill", "tweedie", "mgcv", "broom", "terra"))

library(HAPODASdsmR)
library(Distance)
library(dsm)
library(ggplot2)
library(knitr)
library(tweedie)
library(mgcv)
library(plyr)
library(dplyr)
library(terra)
library(sf)


```

# set up files and objects needed:

## data files:

### hbw DSM Data:

This is the hbw_Dist, hbw_Obs, and hapodas_Segdata, as well as hbw_hr_cds.05 and hbw_hr_mcds_CUE_SIGHTOBS detection function (ddf) objects.

-   hapodas_Segdata: 2km segments(+/-50% since shortes segments merged to previous), no 'off effort' data, Tracklines and transit lines, each segment identified by it's centerpoint.

-   hbw_Dist: all "on effort" observer sightings of hbw and associated sighting covariates used to produce detection funcitons

-   hbw_Obs: all "on effort" observer sightings of hbw with *segment* they are associated with

### Covariate Seg[ment]data

slope, distance to shore, depth (bathymetry), log Depth, and channel type at 2x2km resolution \*\*from ArcGIS: load from from hapodas_segdata

```{r eval=FALSE, include= TRUE}

dataloc <- "~/HAPODASdsmR/data/"  # data location

load(file= paste0(dataloc,"akalbers.rda")) # alaska albers
load(file= paste0(dataloc,"hapodas_Segdata.rda")) # full HBW segment data
load(file= paste0(dataloc,"hbw_Dist.rda")) # full HBW distance data
load(file= paste0(dataloc,"hbw_Obs.rda")) # full HBW observation data
load(file= paste0(dataloc,"hbw_hr_cds.05.rda")) # HBW detection funciton cds
load(file= paste0(dataloc,"hbw_hr_mcds_CUE_SIGHTOBS.rda")) # full HBW detection function mcds
load(file= paste0(dataloc,"SegdataCovs.Rdata")) # full segment covariates


#save(hbw_Dist, hbw_Obs, hapodas_Segdata, hbw_hr_cds.05, hbw_hr_mcds_CUE_SIGHTOBS,file= ppaste0(dataloc,"~/hbwDSMdata2km.Rdata")
```

```{r load functions, eval=FALSE, include= TRUE}
# funloc <- "~/HAPODASdsmR/R/"  # functions location
# 
# load(file= paste0(funloc,"GPTedits_soap_check.R")) # soap_check
# load(file= paste0(funloc,"GPTedits_autocruncher.R")) # autocruncher
# #load(file= paste0(funloc,"___.R")) # ____

```

Because the sightings were truncated by the detection funciton, the observation data table needs to truncate the same sightings:

```{r truncate-obs, eval=FALSE, include= TRUE}
hbw_Obs_t <- hbw_Obs[hbw_Obs$distance <= hbw_hr_mcds_CUE_SIGHTOBS$ddf$meta.data$width, ]


```

### Covariate Seg[ment]data

slope, distance to shore, depth (bathymetry), log Depth, and channel type at 2x2km resolution \*\*from ArcGIS: load from from hapodas_segdata


### hbw DSM data into sf objects
 with akalbers projection

```{r, eval=FALSE, include= TRUE}

# create sf object from projected Alaska Albers coordinates
Segdata_sf <- st_as_sf(hapodas_Segdata,coords = c("x", "y"),# projected coordinates in meters
                       crs = st_crs(akalbers), remove = FALSE)                 

SegdataCovs_sf <- st_as_sf(SegdataCovs,coords = c("x", "y"),# projected coordinates in meters
                       crs = st_crs(akalbers), remove = FALSE)  

#  create sf object from WGS84 lon/lat coordinates
hbw_Obs_t_sf <- st_as_sf(hbw_Obs_t,coords = c("longs", "lats"),# geographic coordinates 
                         crs = 4326, #original CRS = WGS84 
                         remove = FALSE)
      
      #  transform geometry to akalbers
      hbw_Obs_t_sf <- st_transform(hbw_Obs_t_sf,crs = st_crs(akalbers))
      
      #  extract projected coords
      coords <- st_coordinates(hbw_Obs_t_sf)
      
      #  save projected coordinates as x/y columns
      hbw_Obs_t_sf$x <- coords[,1]
      hbw_Obs_t_sf$y <- coords[,2]



```

## shape files:

### survey area/maps

#### Plot survey area and lines, SP transform:

```{r plot survey area and lines, eval=FALSE, include= TRUE}


SEAK <- st_read(                # import coastline shapeobjects
  dsn = "~/HAPODASdsmR/data",
  layer = "Coastal_Land_Area_Dissolve",
  quiet = TRUE
) |>
  #st_set_crs(4326) |>   # assign original CRS FIRST # don't need if have .prj
  st_transform(st_crs(akalbers))


plot(st_geometry(SEAK), axes = TRUE)
   # plot(SEAK, xlab="Northing", ylab="Easting")
    # #lines(Segdata_sf$x, Segdata_sf$y,col='red')
    # points(hbw_Dist$x, hbw_Dist$y, col="blue")

```

### soap film boundaries/surfaces

#### study area (LSF3simpleOCN)

simiplified LSF3ocean shapefile (islands with unserveyed narrow channel between merged)

```{r simplified SEAK, eval=FALSE, include= TRUE}


LSF3ocean <- st_read(
  dsn = "~/HAPODASdsmR/data",
  layer = "new_simpleOCN",   # this is a very simplified outline of the defined study area (used to design the survey) #.shp
  quiet = TRUE
) |>
  st_make_valid() |>
#st_set_crs(4326) |>   # assign original CRS FIRST # don't need if have .prj
   st_transform(st_crs(akalbers))

   save(LSF3ocean,file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/SP_innerSoap.Rda")

 
plot(st_geometry(LSF3ocean), col = "cornflowerblue")
```

#### soap film surface

The soap film surface of the *LSF3surveyArea* to build the knot grid on:

```{r soap film surface for knotgrid: LSF3surveyArea, eval=FALSE, include= TRUE}


coords <- st_coordinates(LSF3ocean) # polygon vertex coords
#head(coords)

coord_list  <- split(
  as.data.frame(coords),
  coords[, "L1"] ) # label by polygon

LSF3surveyArea <- lapply(   # split into lists for soapfilm
  coord_list ,
  function(df) {
    list(
      x = as.numeric(df$X),
      y = as.numeric(df$Y)
    )   }  )

soap_check(bnd = LSF3surveyArea)

```

#### knotgrid

5km knot spacing construction: 'LSF3knots'

```{r define interior LSF3knotskm, eval=FALSE, include= TRUE}
# make grid from extent of observations 


#  extract polygon coordinates from sf geometry
LSF3oceandf <- as.data.frame(st_coordinates(LSF3ocean))

#  rename columns 
names(LSF3oceandf)[names(LSF3oceandf)=="X"] <- "long"
names(LSF3oceandf)[names(LSF3oceandf)=="Y"] <- "lat"

         Nx5 <- floor((max(LSF3oceandf$long)-min(LSF3oceandf$long))/5000) ## 538
         Ny5 <- floor((max(LSF3oceandf$lat)-min(LSF3oceandf$lat))/5000) ## 503

         xknot5 <- seq(min(LSF3oceandf$long),max(LSF3oceandf$long), length.out= Nx5)
         yknot5 <- seq(min(LSF3oceandf$lat),max(LSF3oceandf$lat), length.out= Ny5)

      # setup knots
      LSF3knotgrid <- expand.grid(xknot5,yknot5)
      names(LSF3knotgrid) <- c("x","y")
                     #LSF3knots <- LSF3knotgrid[with(LSF3knotgrid, inSide(bnd = LSF3surveyArea, x, y)), ]## remove knots outside the survey area

       #  convert knot grid to sf points
      knot_sf <- st_as_sf(LSF3knotgrid,
                          coords=c("x","y"),
                          crs=st_crs(LSF3ocean))

      #  identify knots inside survey polygon using sf
      inside <- st_within(knot_sf,
                          st_union(LSF3ocean),
                          sparse=FALSE)[,1]

      #  keep only knots inside survey area
      LSF3knots <- LSF3knotgrid[inside, ]
      
      #save(LSF3knots, file="~/UAF/MS/data and analysis/HAPODAS/LSF3knotskmspacing.Rda")


     # !!!!!!!!!!!load autocruncher.R function first
      soap_check(bnd = LSF3surveyArea, knots=LSF3knots)
      #lines(covSegdata.LSF3soap@coords,col='light blue')
      #points(hbw_obs.LSF3soap@coords, col="white", pch =".")



#load(file="~/UAF/MS/data and analysis/HAPODAS/LSF3knotskmspacing.Rda")

```



#### survey area with 3.75 buffer

3.75km buffer(use with 5km knot spacing)

```{r 3.75km buffer, eval=FALSE, include= TRUE}

      # 
      LSF3_Buff3.75k <- st_read(
         dsn = "~/HAPODASdsmR/data",
         layer = "new2OCN_Buff3.75",   # this is a very simplified outline of the defined study area buffered out by 3.7km
         quiet = TRUE
         ) |>
         st_make_valid() |>
            st_transform(st_crs(akalbers))
            
                 #  extract polygon coordinates from sf geometry
            LSF3coords_buff3.75k <- as.data.frame(
               st_coordinates(LSF3_Buff3.75k)
            )
            
            #  rename coordinate columns 
            names(LSF3coords_buff3.75k)[names(LSF3coords_buff3.75k)=="X"] <- "long"
            names(LSF3coords_buff3.75k)[names(LSF3coords_buff3.75k)=="Y"] <- "lat"
            
            #  
            names(LSF3coords_buff3.75k)[names(LSF3coords_buff3.75k)=="L1"] <- "piece"
      
  

      ###
      
      LSF3borderlist_buff3.75k <- split(LSF3coords_buff3.75k, LSF3coords_buff3.75k$piece)
      names(LSF3borderlist_buff3.75k)
      ###

      
      
       #  keep only coordinate columns
      LSF3SAbuff3.75km <- lapply(
         LSF3borderlist_buff3.75k,
         `[`,  c("long","lat")  )

      #  convert coordinate data.frames into soap-film boundary lists
      nr <- seq_along(LSF3SAbuff3.75km)

      LSF3SAbuff3.75km <- lapply(nr, function(n){

         this_poly <- LSF3SAbuff3.75km[[n]]

         list(
            x = as.numeric(this_poly$long),
            y = as.numeric(this_poly$lat),
            f = rep(0, nrow(this_poly))
         )
      })
      

#load(file="~/UAF/MS/data and analysis/HAPODAS/LSF3buffer.Rda")


```

#### set boundary conditions to zero

```{r, eval=FALSE, include= TRUE}


#  preserve soap boundary structure while adding f vector
Buffer_bnd0 <- lapply(nr, function(n){

   this_poly <- LSF3SAbuff3.75km[[n]]

   #  add boundary condition vector
   this_poly$f <- rep(0, length(this_poly$x))

   return(this_poly)
})


       soap_check(bnd = Buffer_bnd0)
       soap_check(bnd = Buffer_bnd0, knots = LSF3knots)

```
#### find areas w/out boundary set to 0 and replace w/ NA
```{r , eval=FALSE, include= TRUE}


 

outer_buff_bnd_0 <- Buffer_bnd0[[1]]
outer_buff_bnd_0$index <- seq(from =1, by=1, length.out=length(outer_buff_bnd_0$f))

# openings to ocean in order counter-clock-wise from top
A <- data.frame(y=c(1045942,1028017), x= c(1001685,1015511))
B <- data.frame(y=c(854869,833451), x= c(1179819,1207957))
C <- data.frame(y=c(837449,848744), x= c(1222112,1243510))
D <- data.frame(y=c(766392,784364), x= c(1391052,1425770))
E <- data.frame(y=c(790324,796966), x= c(1432990,1449784))


AAA <- which((outer_buff_bnd_0$x <= A$x[2]) & (outer_buff_bnd_0$x >= A$x[1]) &
                (outer_buff_bnd_0$y <= A$y[1]) & (outer_buff_bnd_0$y >= A$y[2]))

BBB <- which((outer_buff_bnd_0$x <= B$x[2]) & (outer_buff_bnd_0$x >= B$x[1]) &
                (outer_buff_bnd_0$y <= B$y[1]) & (outer_buff_bnd_0$y >= B$y[2]))

CCC <- which((outer_buff_bnd_0$x <= C$x[2]) & (outer_buff_bnd_0$x >= C$x[1]) &
                (outer_buff_bnd_0$y <= C$y[2]) & (outer_buff_bnd_0$y >= C$y[1]))

DDD <- which((outer_buff_bnd_0$x <= D$x[2]) & (outer_buff_bnd_0$x >= D$x[1]) &
                (outer_buff_bnd_0$y <= D$y[2]) & (outer_buff_bnd_0$y >= D$y[1]))

EEE <- which((outer_buff_bnd_0$x <= E$x[2]) & (outer_buff_bnd_0$x >= E$x[1]) &
                (outer_buff_bnd_0$y <= E$y[2]) & (outer_buff_bnd_0$y >= E$y[1]))


gaps <- c(AAA,BBB,CCC,DDD,EEE)


for(i in seq_along(outer_buff_bnd_0$f)){
   if (outer_buff_bnd_0$index[i] %in% gaps) {outer_buff_bnd_0$f[i] <- NA}
   else (outer_buff_bnd_0$f[i] <- 0)
}

Buffer_bnd0NA <- Buffer_bnd0

Buffer_bnd0NA[[1]][["f"]] <- outer_buff_bnd_0$f  # New boundary list with boarder conditionas set to 0 everywhere but ocean entrances





```


## Soapfilm adjusted data

Data outside the defined `LSF3surveyArea` must be removed for the soapfilm to work : segdata.LSF3soap

```{r remove segments outside soapfilm, eval=FALSE, include= TRUE}
onoff <- st_within( Segdata_sf,
   st_union(LSF3ocean),sparse = FALSE)[,1]
segdata.LSF3soap <- Segdata_sf[onoff,]
```

```{r remove segments outside soapfilm SegdataCovs_sf, eval=FALSE, include= TRUE}
onoff <- st_within( SegdataCovs_sf,
   st_union(LSF3ocean),sparse = FALSE)[,1]
covSegdata.LSF3soap <- SegdataCovs_sf[onoff,]
```

### hbw_obs.LSF3soap

```{r remove observations outside soapfilm, eval=FALSE, include= TRUE}
onoff <- st_within( hbw_Obs_t_sf,
   st_union(LSF3ocean),sparse = FALSE)[,1]
hbw_obs.LSF3soap <- hbw_Obs_t_sf[onoff,] ## 373 observations
```


# =========

# simple DSMs (no soap)

## x & y

```{r xy_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_nb_xy <- dsm(abundance.est~s(x, y, k=50),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_xy)
###xy_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis_gam(dsm_nb_xy, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nb_xy, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r xy_nb_k25cds, eval=FALSE, include= TRUE}
dsm_nb_xy_cds <- dsm(abundance.est~s(x, y, k=25),
                 ddf.obj=hbw_hr_cds.05, segment.data = hapodas_Segdata,
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_xy_cds)

vis_gam(dsm_nb_xy_cds, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r x+y_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_x_y <- dsm(abundance.est~s(x)+s(y),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata,
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_x_y)

vis_gam(dsm_nb_x_y, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r compare: x+y_nb_mcds v xy_nb_k50mcds, fig.width=12, fig.height=5, eval=FALSE, include= TRUE}
par(mfrow=c(1,2))
vis_gam(dsm_nb_xy, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y, k=50) (link scale)", asp=1)
vis_gam(dsm_nb_x_y, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x) + s(y) (link scale)", asp=1)
```

```{r xy_nb_k200mcds, eval=FALSE, include= TRUE}
tictoc::tic()
dsm_nb_xyK <- dsm(abundance.est~s(x, y, k=200),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata,
                 observation.data=hbw_Obs_t,
                 family=nb())
tictoc::toc()

summary(dsm_nb_xyK)

vis_gam(dsm_nb_xyK, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r compare: xy_nb_k50mcds v xy_nb_k200mcds, fig.width=12, fig.height=5, eval=FALSE, include= TRUE}
par(mfrow=c(1,2))
vis_gam(dsm_nb_xy, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y, k=50) (link scale)", asp=1)
vis_gam(dsm_nb_xyK, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y,k=200 (link scale)", asp=1)
```

```{r xy_tw_k50mcds, eval=FALSE, include= TRUE}
tictoc::tic()
dsm_tw_xy <- dsm(abundance.est~s(x, y, k=50),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata,
                 observation.data=hbw_Obs_t,
                 family=tw())
tictoc::toc()

dsm_tw_xyK <- dsm(abundance.est~s(x, y, k=200),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata,
                 observation.data=hbw_Obs_t,
                 family=tw())

summary(dsm_tw_xy)
par(mfrow=c(2,2))
vis_gam(dsm_nb_xy, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="nb:s(x,y, k=50) (link scale)", asp=1)
vis_gam(dsm_nb_xyK, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y,k=200 (link scale)", asp=1)
vis_gam(dsm_tw_xy, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="tw:s(x,y, k=50) (link scale)", asp=1)
vis_gam(dsm_tw_xyK, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="tw:s(x,y, k=200) (link scale)", asp=1)
```

### summary and interpretation

```{r summarize xy simple models, eval=FALSE, include= TRUE}
# function to extract various bits of information from
# a model object, you can adapt to be more informative
summarize_dsm <- function(model, model_name){
   summ <- summary(model)
         # extract k
         k_vals <- tryCatch(sapply(model$smooth, function(x) x$bs.dim),error = function(e) NA)
   data.frame(response = model$family$family,
              terms    = paste(rownames(summ$s.table), collapse=", "),
              k        = paste(k_vals, collapse=", "),
              AIC      = AIC(model),
              edf      = sum(summ[["edf"]]),
              DevExp   = (summ[["dev.expl"]]*100),
              deviance = model[["deviance"]],
              pvalue  = summ[["p.pv"]][["(Intercept)"]]
   )
}

```

We can then make a table, in this case sorting by AIC *(though note that we can't compare between Tweedie and negative binomial models as the former is a continuous distribution and the latter isn't)*:

```{r sort xy simple models-table, eval=FALSE, include= TRUE}
 # make a named list of models, in this case just the count models
model_list <- mget(c("dsm_nb_xy", "dsm_nb_x_y", "dsm_tw_xy", "dsm_nb_xyK", "dsm_tw_xyK"))
# run this over each model in the list and make a table
summary_table <- do.call(rbind, Map(summarize_dsm, model_list,names(model_list)))
#  sort models from lowest (best) to highest AIC
summary_table <- summary_table[order(summary_table$AIC),]
summary_table
```



```{r check fit-xy simple models, eval=FALSE, include= TRUE}

par(mfrow=c(2,2))
qq.gam(dsm_nb_xy, main="nb:s(x,y, k=25) (link scale)", asp=1, rep=100)
qq.gam(dsm_nb_xyK, main="nb:s(x,y, k=200) (link scale)", asp=1, rep=100)
qq.gam(dsm_nb_xy, main="tw:s(x,y,k=25) (link scale)", asp=1, rep=100)
qq.gam(dsm_nb_xyK, main="tw:s(x,y,k=200) (link scale)", asp=1, rep=100)
```

```{r check fit nb v tw, eval=FALSE, include= TRUE}

par(mfrow=c(1,2))
qq.gam(dsm_nb_xy, main="s(x,y, k=25) (nb link scale)", asp=1, rep=1000)
qq.gam(dsm_tw_xy, main="s(x,y, k=25) (tw link scale)", asp=1, rep=1000)
```


## covariates only

### single covariate

```{r depth_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_depth <- dsm(abundance.est~s(depth),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_depth)
###depth_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis_gam(dsm_nb_depth, view=c("depth"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_depth)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r slope_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_slope <- dsm(abundance.est~s(slope),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_slope)
###slope_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis_gam(dsm_nb_slope, view=c("slope"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_slope)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r d2s_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_d2s <- dsm(abundance.est~s(d2s),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb()) ## d2s = distance to shore

summary(dsm_nb_d2s)
###d2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis_gam(dsm_nb_d2s, view=c("d2s"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_d2s)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

### 2 covariates

```{r depth_slope_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_depth_slope <- dsm(abundance.est ~ s(depth) + s(slope),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_depth_slope)
###depth_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis_gam(dsm_nb_depth_slope, view=c("depth"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_depth_slope)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r d2s_slope_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_d2s_slope <- dsm(abundance.est ~ s(d2s) + s(slope),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_d2s_slope)
###d2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis_gam(dsm_nb_d2s_slope, view=c("d2s"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_d2s_slope)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r depth_d2s_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_depth_d2s <- dsm(abundance.est ~ s(depth) + s(d2s),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_depth_d2s)
###depth_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis_gam(dsm_nb_depth_d2s, view=c("depth"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_depth_d2s)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

## x, y, & covariates

```{r xy_depth_slope_d2s_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_nb_xy_depth_slope_d2s <- dsm(abundance.est ~ s(x, y, k=50)+
                                    s(log(DEPTH)) + s(slope) + s(log(d2s)),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_xy_depth_slope_d2s)
###xy_depth_slope_d2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis_gam(dsm_nb_xy_depth_slope_d2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nb_xy_depth_slope_d2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r xy_depth_slope_d2s_nb_ts, eval=FALSE, include= TRUE}
dsm_nbts_xy_depth_slope_d2s <- dsm(abundance.est ~ s(x, y, bs="ts")+
                                    s(depth, bs="ts") + 
                                      s(slope, bs="ts") + 
                                      s(d2s, bs="ts"),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nbts_xy_depth_slope_d2s)
###xy_depth_slope_d2s_nbts_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis_gam(dsm_nbts_xy_depth_slope_d2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nbts_xy_depth_slope_d2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

## no soap +covs

```{r dsm_xy_S.D.D2s_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_xy_S.LD.LD2s <- dsm(abundance.est ~ s(x, y, k=50)+
                                    s(log(DEPTH)) + s(log(slope)) + s(log(d2s)),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = SegdataCovs, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_xy_S.LD.LD2s)
###xy_logD.D2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis_gam(dsm_xy_S.LD.LD2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_xy_S.LD.LD2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

### + log covs

```{r xy_logD.D2s_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_xy_logD.D2s <- dsm(abundance.est ~ s(x, y, k=50)+
                                    s(log(DEPTH)) + s(slope) + s(log(d2s)),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = SegdataCovs, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_xy_logD.D2s)
###xy_logD.D2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis_gam(dsm__xy_logD.D2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_xy_logD.D2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r xy_logD.D2s_nb_ts, eval=FALSE, include= TRUE}
dsm_nbts_xy_logD.D2s <- dsm(abundance.est ~ s(x, y, bs="ts")+
                                    s(depth, bs="ts") + 
                                      s(slope, bs="ts") + 
                                      s(d2s, bs="ts"),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = SegdataCovs, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nbts_xy_logD.D2s)
###xy_logD.D2s_nbts_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis_gam(dsm_nbts_xy_logD.D2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nbts_xy_logD.D2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

# =========

# Soap DSMs

naming format: covs_S_knotgridspacing#\_buffersize#\_k_nmax \## x & y \#### 5km knots 3.75km buffer 150nmax :

```{r dsm: xysoapLSF3, eval=FALSE, include= TRUE}
tictoc::tic()
xysoapLSF3 <- dsm(abundance.est~s(x, y, bs="so", k=4,
                                      xt=list(bnd=LSF3SAbuff3.75km, nmax=150)),
                 ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, segment.data = segdata.LSF3soap,
                 observation.data=hbw_obs.LSF3soap, knots=LSF3knots,##
                 family=tw(), method="REML")
tictoc::toc()
```

```{r save/load: xysoapLSF3, eval=FALSE, include= TRUE}
 #save(xysoapLSF3, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xysoapLSF3.RData")
#
load( file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xysoapLSF3.RData")
```

```{r summary: xysoapLSF3 , eval=FALSE, include= TRUE }
summary(xysoapLSF3)
```

```{r vis_gam plot: xysoapLSF3, eval=FALSE, include= TRUE}
#par(mfrow=c(1,2))
vis_gam(xysoapLSF3, view=c("x","y"), plot.type="contour", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)

vis_gam(xysoapLSF3, view=c("x","y"), plot.type="contour", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100,
        too.far = )

vis_gam(xysoapLSF3, view=c("x","y"), plot.type="persp", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
vis_gam(xysoapLSF3, view=c("x","y"), plot.type="persp", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
```

```{r plot: xysoapLSF3, eval=FALSE, include= TRUE}
plot(xysoapLSF3)


```

```{r gam.check: xysoapLSF3, eval=FALSE, include= TRUE}
gam.check(xysoapLSF3)

```

```{r check fit :xysoapLSF3, eval=FALSE, include= TRUE}

qq.gam(xysoapLSF3, main="xysoapLSF3", asp=1, rep=100)

rqgam.check(xysoapLSF3)
```

```{r prediction attempt 1: xysoapLSF3, eval=FALSE, include= TRUE}

xysoapLSF3_pred1 <- predict(xysoapLSF3, newdata = LSF3knots, off.set = rep(gridarea,length(knots5$x)))

```

```{r uncerrtainty xysoapLSF3, eval=FALSE, include= TRUE}

dsm.var.gam(xysoapLSF3, pred.data = # ? #
               )


```

## x, y, & covariates

### single covariate

```{r xySdepth, eval=FALSE, include= TRUE}
xySdepth <- dsm(abundance.est ~ s(x, y, bs="so", k=5,
                                      xt=list(bnd=LSF3SAbuff3.75km, nmax=150))+
                                          s(DEPTH, bs="ts"),
                 ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = covSegdata.LSF3soap,
                 observation.data=hbw_obs.LSF3soap ,
                 knots=LSF3knots,##
                 family=tw(), method="REML")


```

```{r single cov summary, eval=FALSE, include= TRUE}
#xySdepth <- xysoap_5_3.75_5_150

#save(xySdepth, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySdepth.RData")

load( "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySdepth.RData")

summary(xySdepth)

plot(xySdepth)

```

```{r xySlogD, eval=FALSE, include= TRUE}
xySlogD <- dsm(abundance.est ~ s(x, y, bs="so", k=5,
                                      xt=list(bnd=LSF3SAbuff3.75km, nmax=150))+
                                          s(log(DEPTH), bs="ts"),
                 ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = covSegdata.LSF3soap,
                 observation.data=hbw_obs.LSF3soap , 
                 knots=LSF3knots,##
                 family=tw(), method="REML")


```

```{r xySlogD single cov summary, eval=FALSE, include= TRUE}

save(xySlogD, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogD.RData")

load( "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogD.RData")

summary(xySlogD)

plot(xySlogD)

```

### 2 covariates

```{r save full data, eval=FALSE, include= TRUE}
save(hbw_Dist, hbw_Obs, hapodas_Segdata, hbw_hr_cds.05, hbw_hr_mcds_CUE_SIGHTOBS,file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/hbwDSMdata2km.Rdata")

```

### 3 covariates

```{r dsm: xySlogd2sD, eval=FALSE, include= TRUE}
tictoc::tic()
xySlogd2sD <- dsm(abundance.est~s(x, y, bs="so", k=4,
                               xt=list(bnd=LSF3SAbuff3.75km, nmax=150))+
                s(log(DEPTH)) + 
                s(slope) + 
                s(log(d2s)),
                ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, 
                segment.data = covSegdata.LSF3soap,
                observation.data=hbw_obs.LSF3soap, 
                knots=LSF3knots,##
                 family=tw(), method="REML")
tictoc::toc()
```

```{r save/load: xySlogd2sD, eval=FALSE, include= TRUE}
 #save(xySlogd2sD, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogd2sD.RData")
#
load( file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogd2sD.RData")
```

```{r summary: xySlogd2sD , eval=FALSE, include= TRUE }
summary(xySlogd2sD)
```

```{r vis_gam plot: xySlogd2sD, eval=FALSE, include= TRUE}
#par(mfrow=c(1,2))
vis_gam(xySlogd2sD, view=c("x","y"), plot.type="contour", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)

vis_gam(xySlogd2sD, view=c("x","y"), plot.type="contour", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100,
        too.far = )

vis_gam(xySlogd2sD, view=c("x","y"), plot.type="persp", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
vis_gam(xySlogd2sD, view=c("x","y"), plot.type="persp", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
```

```{r plot xySlogd2sD, eval=FALSE, include= TRUE}
plot(xySlogd2sD)

```

# =========
