---
title: "hbw_DF_DSM"
author: "Abigail Schiffmiller"
date: "2025-06-23"
output:
  pdf_document: default
  html_document: default
---

## libraries

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is based off the Dolphin dsm example from the distance sampling website: [`http://examples.distancesampling.org/dsm-line-dolphins/mexico-analysis.html`]

```{r libraries , eval=FALSE, include= TRUE}
# install.packages(c("dsm", "Distance", "knitr", "distill", "ggplot2", "rgdal",
#                    "maptools", "plyr", "tweedie"))

#library(HAPODASdsmR)
library(Distance)
library(dsm)
library(ggplot2)
library(knitr)
library(rgdal)
library(maptools)
library(tweedie)
library(mgcv)
library(rgeos)
library(broom)
library(plyr)
library(dplyr)


```

# set up files and objects needed:

## data files:

### hbw DSM Data:

This is the hbw_Dist, hbw_Obs, and hapodas_Segdata, as well as hbw_hr_cds.05 and hbw_hr_mcds_CUE_SIGHTOBS detection function (ddf) objects.

-   hapodas_Segdata: 2km segments(+/-50% since shortes segments merged to previous), no 'off effort' data, Tracklines and transit lines, each segment identified by it's centerpoint.

-   hbw_Dist: all "on effort" observer sightings of hbw and associated sighting covariates used to produce detection funcitons

-   hbw_Obs: all "on effort" observer sightings of hbw with *segment* they are associated with

```{r hbw DSM data, eval=FALSE, include= TRUE}
load("~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/hbwDSMdata2km.Rdata")

#save(hbw_Dist, hbw_Obs, hapodas_Segdata, hbw_hr_cds.05, hbw_hr_mcds_CUE_SIGHTOBS,file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/hbwDSMdata2km.Rdata")
```

```{r load akalbers, eval=FALSE, include= TRUE}
#load("~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/hbwDSMdata2km.Rdata")
load("~/UAF/MS/data and analysis/HAPODAS/HAPODASdsmR/data/akalbers.rda")
```

Because the sightings were truncated by the detection funciton, the observation data table needs to truncate the same sightings:

```{r truncate-obs, eval=FALSE, include= TRUE}
hbw_Obs_t <- hbw_Obs[hbw_Obs$distance <= hbw_hr_mcds_CUE_SIGHTOBS$ddf$meta.data$width, ]

```

### Covariate Seg[ment]data

slope, distance to shore, depth (bathymetry), log Depth, and channel type at 2x2km resolution \*\*from ArcGIS: load from from hapodas_segdata

```{r load Segdata with covariates **from ArcGIS, eval=FALSE, include= TRUE}
 #=================     
      # SegdataCovs <- data.frame(hapodas_Segdata,
      #                          slope = SegdataCovs$slope,
      #                          d2s = SegdataCovs$d2s,
      #                          DEPTH = SegdataCovs$DEPTH,
      #                          logDEPTH = SegdataCovs$logDEPTH,
      #                          ChannelType = SegdataChannel$ChanelType)
 

#  save(SegdataCovs, file="~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/SegdataCovs.Rdata" )

load( file="~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/SegdataCovs.Rdata" )

```

### SPATIAL hbw DSM data

```{r Seg/covSeg/Obs SP, eval=FALSE, include= TRUE}
segSP <- makeSP(dataset = hapodas_Segdata,
                coordx = hapodas_Segdata$longitude,
                coordy = hapodas_Segdata$latitude)

                 
covSegSP <- makeSP(dataset = SegdataCovs,
                coordx = SegdataCovs$longitude,
                coordy = SegdataCovs$latitude)

hbw_obsSP <- makeSP(dataset=hbw_Obs_t,
                coordx = hbw_Obs_t$longs,
                coordy = hbw_Obs_t$lats)



```

## shape files:

### survey area/maps

#### Plot survey area and lines, SP transform:

```{r plot survey area and lines, eval=FALSE, include= TRUE}
SEAK<-readOGR(dsn="C:/Users/aschi/Dropbox/My PC (DESKTOP-3OIUQ56)/Documents/UAF/MS/data and analysis/maps+spatial data/HAPODAS GIS/GISFILES",layer="Coastal_Land_Area_Dissolve") # import coastline shapeobjects

SEAK <- spTransform(SEAK, akalbers)

    plot(SEAK, xlab="Northing", ylab="Easting")
    # #lines(segSP$x, segSP$y,col='red')
    # points(hbw_Dist$x, hbw_Dist$y, col="blue")

```

### soap film boundaries/surfaces

#### study area (LSF3simpleOCN)

simiplified LSF3ocean shapefile (islands with unserveyed narrow channel between merged)

```{r simplified SEAK, eval=FALSE, include= TRUE}
LSF3simpleOCN<- readOGR(dsn="C:/Users/aschi/Dropbox/My PC (DESKTOP-3OIUQ56)/Documents/UAF/MS/data and analysis/maps+spatial data/HAPODAS GIS/HAPODAS shape files/SEAK simplifications",layer="new_simpleOCN") # this is a very simplified outline of the defined study area (used to design the survey)
LSF3ocean<- spTransform(LSF3simpleOCN, CRSobj = akalbers)

which(rgeos::gIsValid(LSF3ocean)==F)

   save(LSF3ocean,file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/SP_innerSoap.Rda")

# par(mar=c(0,0,0,0))
# plot(LSF3ocean, col="cornflowerblue")

```

#### soap film surface

The soap film surface of the *LSF3surveyArea* to build the knot grid on:

```{r soap film surface for knotgrid: LSF3surveyArea, eval=FALSE, include= TRUE}

   LSF3oceandf <- broom::tidy(LSF3ocean)

   ###
   LSF3oceancoords <- LSF3oceandf %>% dplyr::select(long,lat,piece)
   names(LSF3oceancoords) <- c("x", "y", "piece")
   borderlist <- split(LSF3oceancoords, LSF3oceancoords$piece)
   names(borderlist)
   ###

   LSF3surveyArea <- lapply(borderlist, `[`, c(1,2))
   nr <- seq(1,9)
   ###

   LSF3surveyArea <- lapply(nr, function(n) as.list.data.frame(LSF3surveyArea[[n]]))

   #And then check that the border and knots are in order with the soap_check function

   par(mar=c(0,0,1,0))
   soap_check(bnd = LSF3surveyArea)

   save(LSF3surveyArea, file= "~/UAF/MS/data and analysis/HAPODAS/LSF3surveyArea.rda")


load("~/UAF/MS/data and analysis/HAPODAS/LSF3surveyArea.rda")

```

#### knotgrid

5km knot spacing construction: 'LSF3knots'

```{r define interior LSF3knotskm, eval=FALSE, include= TRUE}
# make grid from extent of observations 

      LSF3oceandf <- broom::tidy(LSF3ocean)

         Nx5 <- floor((max(LSF3oceandf$long)-min(LSF3oceandf$long))/5000) ## 538
         Ny5 <- floor((max(LSF3oceandf$lat)-min(LSF3oceandf$lat))/5000) ## 503

         xknot5 <- seq(min(LSF3oceandf$long),max(LSF3oceandf$long), length.out= Nx5)
         yknot5 <- seq(min(LSF3oceandf$lat),max(LSF3oceandf$lat), length.out= Ny5)

      # setup knots
      LSF3knotgrid <- expand.grid(xknot5,yknot5)
      names(LSF3knotgrid) <- c("x","y")
      LSF3knots <- LSF3knotgrid[with(LSF3knotgrid, inSide(bnd = LSF3surveyArea, x, y)), ]## remove knots outside the survey area

      #save(LSF3knots, file="~/UAF/MS/data and analysis/HAPODAS/LSF3knotskmspacing.Rda")


      par(mar=c(0,0,1,0))
      soap_check(bnd = LSF3surveyArea, knots=LSF3knots)
      lines(covSegdata.LSF3soap@coords,col='light blue')
      #points(hbw_obs.LSF3soap@coords, col="white", pch =".")



load(file="~/UAF/MS/data and analysis/HAPODAS/LSF3knotskmspacing.Rda")

```

#### survey area with 3.75 buffer

3.75km buffer(use with 5km knot spacing)

```{r 3.75km buffer, eval=FALSE, include= TRUE}

      # #
      # 
      LSF3_Buff3.75k <- readOGR(dsn="C:/Users/aschi/Dropbox/My PC (DESKTOP-3OIUQ56)/Documents/UAF/MS/data and analysis/maps+spatial data/HAPODAS GIS/HAPODAS shape files/SEAK buffers",layer="new2OCN_Buff3.75")

      LSF3_Buff3.75k <- spTransform(LSF3_Buff3.75k, CRSobj = akalbers)

      which(rgeos::gIsValid(LSF3_Buff3.75k)==F)

      LSF3df_buff3.75k <- broom::tidy(LSF3_Buff3.75k)
      
   #save(LSF3_Buff3.75k,file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/SP_LSF3buffer.Rda")

      ###
      LSF3coords_buff3.75k <- LSF3df_buff3.75k %>% dplyr::select(long,lat,piece)
      names(LSF3coords_buff3.75k) <- c("x", "y", "piece")
      LSF3borderlist_buff3.75k <- split(LSF3coords_buff3.75k, LSF3coords_buff3.75k$piece)
      names(LSF3borderlist_buff3.75k)
      ###

      LSF3SAbuff3.75km <- lapply(LSF3borderlist_buff3.75k, `[`, c(1,2))
      nr <- seq(1,8)
      ###

      LSF3SAbuff3.75km <- lapply(nr, function(n) as.list.data.frame(LSF3SAbuff3.75km[[n]]))
      # 
      # #And then check that the border and knots are in order with the soap_check function
      # 
      # par(mar=c(0,0,1,0))
      # soap_check(bnd = LSF3SAbuff3.75km)
      # soap_check(bnd = LSF3SAbuff3.75km, knots = LSF3knots)
      # 
      # 
      # #save(LSF3SAbuff3.75km, file="~/UAF/MS/data and analysis/HAPODAS/LSF3buffer.Rda")

load(file="~/UAF/MS/data and analysis/HAPODAS/LSF3buffer.Rda")


```

#### set boundary conditions

```{r set boundary condition to zero, eval=FALSE, include= TRUE}
nr <- seq(1,8)
surveyArea_0 <- lapply(nr, function(n) LSF3SAbuff3.75km[[n]]<- c(LSF3SAbuff3.75km[[n]], list(f = rep(0, length(LSF3SAbuff3.75km[[n]]$x)))))
Buffer_bnd0<- surveyArea_0
      # par(mar=c(0,0,1,0))
      # soap_check(bnd = surveyArea_0)
      # soap_check(bnd = surveyArea_0, knots = LSF3knots)
save(Buffer_bnd0, file = "~/UAF/MS/data and analysis/HAPODAS/Buffer_bnd0.Rda")

```

```{r find areas w/out boundary set to 0 and replace w/ NA, eval=FALSE, include= TRUE}


#Soap_0_buffA1 

fff <- Soap_0_buffA1[[1]][["f"]]
xxx <- Soap_0_buffA1[[1]][["x"]]
yyy <- Soap_0_buffA1[[1]][["y"]]

DF <- data.frame(fff,xxx,yyy)
DF$index <- seq(from =1, by=1, length.out=length(DF$fff))

# openings to ocean in order counter-clock-wise from top
A <- data.frame(y=c(1045942,1028017), x= c(1001685,1015511))
B <- data.frame(y=c(854869,833451), x= c(1179819,1207957))
C <- data.frame(y=c(837449,848744), x= c(1222112,1243510))
D <- data.frame(y=c(766392,784364), x= c(1391052,1425770))
E <- data.frame(y=c(790324,796966), x= c(1432990,1449784))


AAA <- which((DF$xxx <= A$x[2]) & (DF$xxx >= A$x[1]) &
                (DF$yyy <= A$y[1]) & (DF$yyy >= A$y[2]))

BBB <- which((DF$xxx <= B$x[2]) & (DF$xxx >= B$x[1]) &
                (DF$yyy <= B$y[1]) & (DF$yyy >= B$y[2]))

CCC <- which((DF$xxx <= C$x[2]) & (DF$xxx >= C$x[1]) &
                (DF$yyy <= C$y[2]) & (DF$yyy >= C$y[1]))

DDD <- which((DF$xxx <= D$x[2]) & (DF$xxx >= D$x[1]) &
                (DF$yyy <= D$y[2]) & (DF$yyy >= D$y[1]))

EEE <- which((DF$xxx <= E$x[2]) & (DF$xxx >= E$x[1]) &
                (DF$yyy <= E$y[2]) & (DF$yyy >= E$y[1]))


ggg <- c(AAA,BBB,CCC,DDD,EEE)


for(i in seq_along(DF$fff)){
   if (DF$index[i] %in% ggg) {DF$fff[i] <- NA}
   else (DF$fff[i] <- 0)
}

Soap_0_buffA1[[1]][["f"]] <- DF$fff



save(Soap_0_buffA1, file = "~/UAF/MS/data and analysis/HAPODAS/Soap_0_buffA1.Rda")


```

## Soapfilm adjusted data

Data outside the defined `LSF3surveyArea` must be removed for the soapfilm to work \### segdata.LSF3soap

```{r remove segments outside soapfilm, eval=FALSE, include= TRUE}
x <- segSP$x; y <- segSP$y 
onoff <- inSide(x=x,y=y, bnd=as.list(LSF3surveyArea))
rm(x,y)
segdata.LSF3soap <- segSP[onoff,]
```

```{r remove segments outside soapfilm covSegSP, eval=FALSE, include= TRUE}
x <- covSegSP$x; y <- covSegSP$y 
onoff <- inSide(x=x,y=y, bnd=as.list(LSF3surveyArea))
rm(x,y)
covSegdata.LSF3soap <- covSegSP[onoff,]
```

### hbw_obs.LSF3soap

```{r remove observations outside soapfilm, eval=FALSE, include= TRUE}

hbw_obsSP <- spTransform(hbw_obsSP, CRSobj = akalbers) ## 375 observations

x <- hbw_obsSP$x; y <- hbw_obsSP$y
onoff <- inSide(x=x,y=y, bnd=as.list(LSF3surveyArea))
rm(x,y)
hbw_obs.LSF3soap <- hbw_obsSP[onoff,] ## 373 observations
```

```{r LSF3.soapdata, eval=FALSE, include= TRUE}

#save(segdata.LSF3soap, covSegdata.LSF3soap, hbw_obs.LSF3soap, file= "~/UAF/MS/data and analysis/HAPODAS/LSF3soapdata.Rda")

load(file= "~/UAF/MS/data and analysis/HAPODAS/LSF3soapdata.Rda")


```

# =========

# simple DSMs (no soap)

## x & y

```{r xy_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_nb_xy <- dsm(abundance.est~s(x, y, k=50),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_xy)
###xy_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis.gam(dsm_nb_xy, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nb_xy, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r xy_nb_k25cds, eval=FALSE, include= TRUE}
# dsm_nb_xy_cds <- dsm(abundance.est~s(x, y, k=25),
#                  ddf.obj=hbw_hr_cds.05, segment.data = hapodas_Segdata, 
#                  observation.data=hbw_Obs_t,
#                  family=nb())
# 
# summary(dsm_nb_xy_cds)
# 
# vis.gam(dsm_nb_xy_cds, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r x+y_nb_mcds, eval=FALSE, include= TRUE}
# dsm_nb_x_y <- dsm(abundance.est~s(x)+s(y),
#                  ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata, 
#                  observation.data=hbw_Obs_t,
#                  family=nb())
# 
# summary(dsm_nb_x_y)
# 
# vis.gam(dsm_nb_x_y, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r compare: x+y_nb_mcds v xy_nb_k50mcds, fig.width=12, fig.height=5, eval=FALSE, include= TRUE}
# par(mfrow=c(1,2))
# vis.gam(dsm_nb_xy, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x,y, k=25) (link scale)", asp=1)
# vis.gam(dsm_nb_x_y, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x) + s(y) (link scale)", asp=1)
```

```{r xy_nb_k200mcds, eval=FALSE, include= TRUE}
# tictoc::tic()
# dsm_nb_xyK <- dsm(abundance.est~s(x, y, k=200),
#                  ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata, 
#                  observation.data=hbw_Obs_t,
#                  family=nb())
# tictoc::toc()
# 
# summary(dsm_nb_xyK)
# 
# vis.gam(dsm_nb_xyK, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r compare: xy_nb_k50mcds v xy_nb_k200mcds, fig.width=12, fig.height=5, eval=FALSE, include= TRUE}
# par(mfrow=c(1,2))
# vis.gam(dsm_nb_xy, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x,y, k=25) (link scale)", asp=1)
# vis.gam(dsm_nb_xyK, view=c("x","y"), plot.type="contour", 
#         too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

```{r xy_tw_k50mcds, eval=FALSE, include= TRUE}
tictoc::tic()
dsm_tw_xy <- dsm(abundance.est~s(x, y, k=50),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, segment.data = hapodas_Segdata,
                 observation.data=hbw_Obs_t,
                 family=tw())
tictoc::toc()

summary(dsm_tw_xy)

vis.gam(dsm_tw_xy, view=c("x","y"), plot.type="contour",
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
```

### summary and interpretation

```{r summarize xy simple models, eval=FALSE, include= TRUE}
# function to extract various bits of information from
# a model object, you can adapt to be more informative
summarize_dsm <- function(model){
   summ <- summary(model)
   data.frame(response = model$family$family,
              terms    = paste(rownames(summ$s.table), collapse=", "),
              AIC      = AIC(model),
              edf      = summ[["edf"]],
              DevExp   = (summ[["dev.expl"]]*100),
              deviance = model[["deviance"]],
              p-value  = summ[["p.pv"]][["(Intercept)"]]
   )
}
# make a list of models, in this case just the count models
model_list <- list(dsm_nb_xy, dsm_nb_x_y) #, dsm_tw_xy, dsm_tw_x_y
# run this over each model in the list and make a table
library(plyr)
summary_table <- ldply(model_list, summarize_dsm)
row.names(summary_table) <- c("dsm_nb_xy", "dsm_nb_x_y") #, "dsm_tw_xy", "dsm_tw_x_y"
```

We can then again use the `kable` function to make a table, in this case sorting by AIC (though note that we can't compare between Tweedie and negative binomial models as the former is a continuous distribution and the latter isn't):

```{r sort xy simple models-table, eval=FALSE, include= TRUE}
 summary_table <- summary_table[order(summary_table$response),]

 summary_table <- summary_table[order(summary_table$AIC),]
```

```{r print-table, results="asis", eval=FALSE, include= TRUE}
kable(summary_table)
```

```{r check fit-xy simple models, eval=FALSE, include= TRUE}

#par(mfrow=c(1,2))
qq.gam(dsm_nb_xy, main="s(x,y, k=25) (link scale)", asp=1, rep=100)
#qq.gam(dsm_nb_xyK, main="s(x,y) (link scale)", asp=1, rep=100)
```

```{r check fit nb v tw, eval=FALSE, include= TRUE}

par(mfrow=c(1,2))
qq.gam(dsm_nb_xy, main="s(x,y, k=25) (nb link scale)", asp=1, rep=1000)
qq.gam(dsm_tw_xy, main="s(x,y, k=25) (tw link scale)", asp=1, rep=1000)
```

## covariates only

### single covariate

```{r depth_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_depth <- dsm(abundance.est~s(depth),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_depth)
###depth_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis.gam(dsm_nb_depth, view=c("depth"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_depth)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r slope_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_slope <- dsm(abundance.est~s(slope),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_slope)
###slope_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis.gam(dsm_nb_slope, view=c("slope"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_slope)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r d2s_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_d2s <- dsm(abundance.est~s(d2s),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb()) ## d2s = distance to shore

summary(dsm_nb_d2s)
###d2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis.gam(dsm_nb_d2s, view=c("d2s"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_d2s)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

### 2 covariates

```{r depth_slope_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_depth_slope <- dsm(abundance.est ~ s(depth) + s(slope),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_depth_slope)
###depth_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis.gam(dsm_nb_depth_slope, view=c("depth"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_depth_slope)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r d2s_slope_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_d2s_slope <- dsm(abundance.est ~ s(d2s) + s(slope),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_d2s_slope)
###d2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis.gam(dsm_nb_d2s_slope, view=c("d2s"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_d2s_slope)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r depth_d2s_nb_mcds, eval=FALSE, include= TRUE}
dsm_nb_depth_d2s <- dsm(abundance.est ~ s(depth) + s(d2s),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_depth_d2s)
###depth_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
#vis.gam(dsm_nb_depth_d2s, view=c("depth"), plot.type="contour",too.far=0.1, main="s(x,y) (link scale)", asp=1)
#
plot(dsm_nb_depth_d2s)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

## x, y, & covariates

```{r xy_depth_slope_d2s_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_nb_xy_depth_slope_d2s <- dsm(abundance.est ~ s(x, y, k=50)+
                                    s(log(DEPTH)) + s(slope) + s(log(d2s)),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nb_xy_depth_slope_d2s)
###xy_depth_slope_d2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis.gam(dsm_nb_xy_depth_slope_d2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nb_xy_depth_slope_d2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r xy_depth_slope_d2s_nb_ts, eval=FALSE, include= TRUE}
dsm_nbts_xy_depth_slope_d2s <- dsm(abundance.est ~ s(x, y, bs="ts")+
                                    s(depth, bs="ts") + 
                                      s(slope, bs="ts") + 
                                      s(d2s, bs="ts"),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = cov3Segdata, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nbts_xy_depth_slope_d2s)
###xy_depth_slope_d2s_nbts_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis.gam(dsm_nbts_xy_depth_slope_d2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nbts_xy_depth_slope_d2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

## no soap +covs

```{r dsm_xy_S.D.D2s_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_xy_S.LD.LD2s <- dsm(abundance.est ~ s(x, y, k=50)+
                                    s(log(DEPTH)) + s(log(slope)) + s(log(d2s)),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = SegdataCovs, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_xy_S.LD.LD2s)
###xy_logD.D2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis.gam(dsm_xy_S.LD.LD2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_xy_S.LD.LD2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

### + log covs

```{r xy_logD.D2s_nb_k50mcds, eval=FALSE, include= TRUE}
dsm_xy_logD.D2s <- dsm(abundance.est ~ s(x, y, k=50)+
                                    s(log(DEPTH)) + s(slope) + s(log(d2s)),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = SegdataCovs, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_xy_logD.D2s)
###xy_logD.D2s_nb_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis.gam(dsm__xy_logD.D2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_xy_logD.D2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

```{r xy_logD.D2s_nb_ts, eval=FALSE, include= TRUE}
dsm_nbts_xy_logD.D2s <- dsm(abundance.est ~ s(x, y, bs="ts")+
                                    s(depth, bs="ts") + 
                                      s(slope, bs="ts") + 
                                      s(d2s, bs="ts"),
                 ddf.obj=hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = SegdataCovs, 
                 observation.data=hbw_Obs_t,
                 family=nb())

summary(dsm_nbts_xy_logD.D2s)
###xy_logD.D2s_nbts_k50mcds-visgam, fig.width=5, fig.height=5, fig.cap="Fitted surface (on link scale) for s(x,y)"}
vis.gam(dsm_nbts_xy_logD.D2s, view=c("x","y"), plot.type="contour", 
        too.far=0.1, main="s(x,y) (link scale)", asp=1)
plot(dsm_nbts_xy_logD.D2s, scheme=2)
# par(new=TRUE)
# points(hbw_Obs_t$longs,hbw_Obs_t$lats)

```

# =========

# Soap DSMs

naming format: covs_S_knotgridspacing#\_buffersize#\_k_nmax \## x & y \#### 5km knots 3.75km buffer 150nmax :

```{r dsm: xysoapLSF3, eval=FALSE, include= TRUE}
tictoc::tic()
xysoapLSF3 <- dsm(abundance.est~s(x, y, bs="so", k=4,
                                      xt=list(bnd=LSF3SAbuff3.75km, nmax=150)),
                 ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, segment.data = segdata.LSF3soap,
                 observation.data=hbw_obs.LSF3soap, knots=LSF3knots,##
                 family=tw(), method="REML")
tictoc::toc()
```

```{r save/load: xysoapLSF3, eval=FALSE, include= TRUE}
 #save(xysoapLSF3, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xysoapLSF3.RData")
#
load( file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xysoapLSF3.RData")
```

```{r summary: xysoapLSF3 , eval=FALSE, include= TRUE }
summary(xysoapLSF3)
```

```{r vis.gam plot: xysoapLSF3, eval=FALSE, include= TRUE}
#par(mfrow=c(1,2))
vis.gam(xysoapLSF3, view=c("x","y"), plot.type="contour", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)

vis.gam(xysoapLSF3, view=c("x","y"), plot.type="contour", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100,
        too.far = )

vis.gam(xysoapLSF3, view=c("x","y"), plot.type="persp", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
vis.gam(xysoapLSF3, view=c("x","y"), plot.type="persp", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
```

```{r plot: xysoapLSF3, eval=FALSE, include= TRUE}
plot(xysoapLSF3)


```

```{r gam.check: xysoapLSF3, eval=FALSE, include= TRUE}
gam.check(xysoapLSF3)

```

```{r check fit :xysoapLSF3, eval=FALSE, include= TRUE}

qq.gam(xysoapLSF3, main="xysoapLSF3", asp=1, rep=100)

rqgam.check(xysoapLSF3)
```

```{r prediction attempt 1: xysoapLSF3, eval=FALSE, include= TRUE}

xysoapLSF3_pred1 <- predict(xysoapLSF3, newdata = LSF3knots, off.set = rep(gridarea,length(knots5$x)))

```

```{r uncerrtainty xysoapLSF3, eval=FALSE, include= TRUE}

dsm.var.gam(xysoapLSF3, pred.data = # ? #
               )


```

## x, y, & covariates

### single covariate

```{r xySdepth, eval=FALSE, include= TRUE}
xySdepth <- dsm(abundance.est ~ s(x, y, bs="so", k=5,
                                      xt=list(bnd=LSF3SAbuff3.75km, nmax=150))+
                                          s(DEPTH, bs="ts"),
                 ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = covSegdata.LSF3soap,
                 observation.data=hbw_obs.LSF3soap ,
                 knots=LSF3knots,##
                 family=tw(), method="REML")


```

```{r single cov summary, eval=FALSE, include= TRUE}
#xySdepth <- xysoap_5_3.75_5_150

#save(xySdepth, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySdepth.RData")

load( "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySdepth.RData")

summary(xySdepth)

plot(xySdepth)

```

```{r xySlogD, eval=FALSE, include= TRUE}
xySlogD <- dsm(abundance.est ~ s(x, y, bs="so", k=5,
                                      xt=list(bnd=LSF3SAbuff3.75km, nmax=150))+
                                          s(log(DEPTH), bs="ts"),
                 ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, 
                 segment.data = covSegdata.LSF3soap,
                 observation.data=hbw_obs.LSF3soap , 
                 knots=LSF3knots,##
                 family=tw(), method="REML")


```

```{r xySlogD single cov summary, eval=FALSE, include= TRUE}

save(xySlogD, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogD.RData")

load( "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogD.RData")

summary(xySlogD)

plot(xySlogD)

```

### 2 covariates

```{r save full data, eval=FALSE, include= TRUE}
save(hbw_Dist, hbw_Obs, hapodas_Segdata, hbw_hr_cds.05, hbw_hr_mcds_CUE_SIGHTOBS,file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DATA/hbwDSMdata2km.Rdata")

```

### 3 covariates

```{r dsm: xySlogd2sD, eval=FALSE, include= TRUE}
tictoc::tic()
xySlogd2sD <- dsm(abundance.est~s(x, y, bs="so", k=4,
                               xt=list(bnd=LSF3SAbuff3.75km, nmax=150))+
                s(log(DEPTH)) + 
                s(slope) + 
                s(log(d2s)),
                ddf.obj = hbw_hr_mcds_CUE_SIGHTOBS, 
                segment.data = covSegdata.LSF3soap,
                observation.data=hbw_obs.LSF3soap, 
                knots=LSF3knots,##
                 family=tw(), method="REML")
tictoc::toc()
```

```{r save/load: xySlogd2sD, eval=FALSE, include= TRUE}
 #save(xySlogd2sD, file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogd2sD.RData")
#
load( file= "~/UAF/MS/data and analysis/HAPODAS/HAPODAS DSMs/xySlogd2sD.RData")
```

```{r summary: xySlogd2sD , eval=FALSE, include= TRUE }
summary(xySlogd2sD)
```

```{r vis.gam plot: xySlogd2sD, eval=FALSE, include= TRUE}
#par(mfrow=c(1,2))
vis.gam(xySlogd2sD, view=c("x","y"), plot.type="contour", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)

vis.gam(xySlogd2sD, view=c("x","y"), plot.type="contour", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100,
        too.far = )

vis.gam(xySlogd2sD, view=c("x","y"), plot.type="persp", type="link", 
        main="s(x,y) (link scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
vis.gam(xySlogd2sD, view=c("x","y"), plot.type="persp", type="response", 
        main="s(x,y) (response scale)",n.grid = 95, asp=1, color = "heat", nCol = 100)
```

```{r plot xySlogd2sD, eval=FALSE, include= TRUE}
plot(xySlogd2sD)

```

# =========
